Generalized Tractability

### 2023-03-11 GeneralizedTractability_230310.tex
Fred M2 2023 Mac Updates

@@

\subsection{Notation and fundamental definitions}
Let $T$ be a tractability function defined as follows,
\[
T :(0,\infty) \times \mathbb{N} \times [0,\infty)^s \rightarrow [1,\infty),
\]
A tractability function is a simple function that provides an upper bound on the computational complexity of a problem
A problem is $T$-tractable with parameter $\vp$ iff there exists a positive constant $C_{\vp}$ such that
\begin{equation} \label{eq:tract_def}
	\comp(\varepsilon,d) \le C_{\vp}\, T(\varepsilon^{-1},d,\vp) \qquad \forall \varepsilon >0, \ d \in \mathbb{N}.
\end{equation}
A problem is \emph{strongly}
$T$-tractable with parameter $\vp$ iff the complexity is independent of the dimension, that is, there exists a positive constant $C_{\vp}$ such that
\begin{equation} \label{eq:tract_def}
	\comp(\varepsilon,d) \le C_{\vp}\, T(\varepsilon^{-1},1,\vp) \qquad \forall \varepsilon >0 , \ d \in \mathbb{N}.
\end{equation}

@@

If \eqref{eq:tract_def} holds for some $\vp$, it clearly holds for larger, $\vp$.  We are often interested in the optimal or smallest $\vp$ for which \eqref{eq:tract_def} holds.
The  set of optimal parameters, $\mathcal{P}_{\text{opt}}$ is defined as
\begin{equation}
	\mathcal{P}_{\textup{opt}} : = \{\vp^* : \eqref{eq:tract_def} \text{ holds }\forall \vp \in [\vp,\boldsymbol{\infty}) \setminus \{\vp^*\} \text{ and }
	\eqref{eq:tract_def} \text{ fails } \forall \vp \in [\vzero,\vp] \setminus \{\vp^*\} \}.
\end{equation}
In this article we allow $\vp$ to be a scalar or vector as the situation requires.

@@

In the sections below we prove necessary and sufficient conditions on tractability as generally defined in \eqref{eq:tract_def}.  These conditions  that involve the boundedness of sums defined in terms of $\{T(\lambda_{i,d}^{-1},\cdot, \cdot)\}_{i=1}^\infty$.  In practice, it may be easier to verify whether or not these conditions hold than to deal with \eqref{eq:comp_def} directly.

@@

\begin{theorem}\label{thm_main_strong_tract2}
A problem is strongly $T$-tractable iff there exists $\vp \in [\vzero, \boldsymbol{\infty})$ and an integer $L(\vp) > 0$ such that
\begin{equation} \label{eq:strong_tractiff3}
     S(p):=\sup_{d \in \naturals} \sum_{i = L(\vp)}^\infty \frac{1}{T(\lambda_{i,d}^{-1},1,\vp)} < \infty.
\end{equation}
If \eqref{eq:strong_tractiff3} holds for some $\vp$, then
\[
	\widetilde{\mathcal{P}}_{\textup{opt}} := \{\vp^* : \eqref{eq:strong_tractiff3} \text{ holds }\forall \vp \in [\vp,\boldsymbol{\infty}) \setminus \{\vp^*\} \text{ and }
\eqref{eq:strong_tractiff3} \text{ fails } \forall \vp \in [\vzero,\vp] \setminus \{\vp^*\} \} = \mathcal{P}}_{\textup{opt}.
\]
\end{theorem}

\begin{proof}
\textbf{Sufficient condition:}\\
Given that $T$ is increasing in its arguments, we have the following equivalent expression for the complexity:
    \begin{equation*}
    \comp(\varepsilon, d) &= \min \{n \in \natzero : \lambda_{n+1, d} \le \varepsilon\} = \min \biggl\{n \in \natzero : \frac{1}{T(\lambda_{n+1,d}^{-1},1,\vp)}\le \frac{1}{T(\varepsilon^{-1},1,\vp) }\biggr\}.
\end{equation*}
Also since the $\lambda_{i,d}$ are non-increasing, it follows that the $T(\lambda_{i,d}^{-1},1,\vp)$ are non-decreasing. In particular, for all $n\in \mathbb{N}$, we have

\begin{align*}
    \MoveEqLeft{\lambda_{n+1,d} \le \lambda_{n,d} \le \cdots \le \lambda_{1,d}} \\
    & \implies \frac{1}{T(\lambda_{n+1, d}^{-1},1,\vp)} \le \frac{1}{T(\lambda_{n, d}^{-1},1,\vp)} \le \cdots \le \frac{1}{T(\lambda_{1, d}^{-1},1,\vp)} \\
    & \implies \frac{1}{T(\lambda_{n+1, d}^{-1},1,\vp) }
    \le \frac 1n \sum_{i=1}^n  \frac{1}{T(\lambda_{i, d}^{-1},1,\vp) }
    \le \frac 1n \sum_{i=1}^\infty  \frac{1}{T(\lambda_{i, d}^{-1},1,\vp)}.
\end{align*}
Thus, we can conclude that
\begin{equation*}
    n \ge T(\varepsilon^{-1},1,\vp) \sum_{i=1}^\infty \frac{1}{T(\lambda_{i, d}^{-1},1,\vp)}
   \implies   \frac{1}{T(\lambda_{n+1, d}^{-1},1,\vp)} \le
   \frac 1n \sum_{i=1}^\infty \frac{1}{T(\lambda_{i, d}^{-1},1,\vp) } \le \frac{1}{T(\varepsilon^{-1},1,\vp)}.
\end{equation*}

This inequality implies an upper bound on the complexity of
\begin{align*}
       \comp(\varepsilon,d)
       &\le 1 + T(\varepsilon^{-1},1,\vp) \sum_{i=1}^\infty \frac{1}{T(\lambda_{i, d}^{-1},1,\vp)} \\
       & = 1 + T(\varepsilon^{-1},1, \vp) \left [ \sum_{i=1}^{L(p)-1} \frac{1}{T(\lambda_{i, d}^{-1},1,\vp)}
       + \underbrace{\sum_{i=L(p)}^\infty \frac{1}{T(\lambda_{i, d}^{-1},1,\vp)}}_{\le S(p) <\infty\ \text{by \eqref{eq:strong_tractiff3}} }\right] \\
       & \le 1+ T(\varepsilon^{-1},1,\vp) \underbrace{\left[\frac{(L(p)-1)}{T(\lambda_{\max}^{-1},1,\vp)} + S(p)\right]}_{=:C_1(p)}\\
       & = 1+ T(\varepsilon^{-1},1,\vp) C_1 (p)\\
       &\le C_{\vp} T(\varepsilon^{-1},1,\vp),
\end{align*}
for some suitably chosen $C_{\vp}>0$. This means that we have strong $T$-tractability, and verifies sufficiency of \eqref{eq:strong_tractiff3}.

\bigskip
\noindent \textbf{Necessary condition:} \\
Suppose that we have strong
$T$-tractability. That is, for some $\vp \ge \vzero$ there exists a positive constant $C_{\vp}$ such that
\[
\comp(\varepsilon,d)\le C_{\vp}\, T(\varepsilon^{-1},1,\vp)
\qquad \forall \varepsilon > 0, \ d \in \mathbb{N}.
\]
Since the eigenvalues $\lambda_{j,d}$ are non-increasing, we have
\begin{equation}\label{eq:lambda_K_strong3}
\lambda_{\lfloor C_{\vp}\, T(\varepsilon^{-1},1,\vp)\rfloor +1,d}\le \varepsilon.
\end{equation}

@@

For $\varepsilon$ taking on all positive values, $j(\varepsilon,\vp)$ takes on all values greater than or equal to $L(\vp)$, so
\[
j\le C_{\vp}\, T(\varepsilon^{-1},1,\vp)+1 \le C_{\vp} T(\lambda_{j,d}^{-1},1,\vp)+1, \qquad \forall j \ge L(\vp).
\]
This implies via our technical assumption \eqref{eq:ptauassume} that
\begin{equation*}
 K_{\vp,\tau}\,T (\lambda_{j,d}^{-1},1,\tau \vp) \ge
 [T(\lambda_{j,d}^{-1},1, \vp)]^\tau
 \ge
  \left[\frac{(j-1)}{C_{\vp}}\right]^\tau \qquad \forall j \ge L(\vp)
\end{equation*}
Summing both sides of the inequality yields and noting that $L(\vp) \le L(\tau\v)$, it follows that
\begin{multline*}
\sup_{d\in\naturals} \sum_{j=L(\tau\vp)}^\infty \frac{1}{T (\lambda_{j,d}^{-1},1, \tau \vp)} \le
\sup_{d\in\naturals} \sum_{j=L(\vp)}^\infty \frac{1}{T (\lambda_{j,d}^{-1},1, \tau \vp)} \\
\le K_{\vp,\tau}\, C_{\vp}^\tau
\sum_{j=L(\vp)}^\infty \frac{1}{(j-1)^\tau} \le
K_{\vp,\tau}\, C_{\vp}^\tau\, \zeta (\tau) < \infty,
\end{multline*}
where $\zeta$ denotes the Riemann zeta function.
This yields \eqref{eq:strong_tractiff3} with $p$ replaced by $\tau p$, and so we see the necessity of $\eqref{eq:strong_tractiff3}$.

\bigskip
\noindent \textbf{Optimality:} \\
To complete the proof, we must show that $\widetilde{\cp}_{\textup{opt}} = \cp_{\textup{opt}}$.

Assume that $\mathcal{P}_{\text{opt}}=(p^*,0)$. We already know from the first part of the proof that
\[
 p^* \le \inf\{p>0\ \mbox{such that \eqref{eq:strong_tractiff3} holds}\}.
\]
Suppose that we would have
\[
 p^* < \inf\{p>0\ \mbox{such that \eqref{eq:strong_tractiff3} holds}\}.
\]
\peter{Then, however, there exists a $p^\dag$ such that
\[
p^* < p^\dag < \inf\{p>0\ \mbox{such that \eqref{eq:strong_tractiff3} holds}\},
\]
and that the problem is $T$-tractable with parameters $q=0$ and $p^\dag$. }
We can proceed as in the second part of the proof and show that
\[
 \sup_{d\in\naturals} \sum_{j=L(\tau \peter{p^\dag})}^\infty \frac{1}{T(\lambda_{j,d}^{-1},1, \tau \peter{p^\dag},0)}<\infty.
\]
for any $\tau>1$, and a suitable $L(\tau \peter{p^\dag})$. By choosing $\tau$ sufficiently close to 1, we then would
obtain
\[
 \tau \peter{p^\dag} < \inf\{p>0\ \mbox{such that \eqref{eq:strong_tractiff3} holds}\},
\]
which is a contradiction. So, we must have
\[
 p^* = \inf\{p>0\ \mbox{such that \eqref{eq:strong_tractiff3} holds}\}.
\]

This concludes the proof of the theorem.

As for $\mathcal{P}_{\text{opt}}$, we immediately see from the above argument that $\mathcal{P}_{\text{opt}}=(p^*,0)$, where
\[
p^* \le \inf\{p>0\ \mbox{such that \eqref{eq:strong_tractiff3} holds}\}.
\]

>>>>

\newcommand{\vinfty}{\boldsymbol{\infty}}

@@

\subsection{Notation and fundamental definitions}
Let $T$ be a tractability function defined as follows,
\[
T :(0,\infty) \times \mathbb{N} \times [0,\infty)^s \rightarrow [1,\infty),
\]
A tractability function is a simple function that provides an upper bound on the computational complexity of a problem
A problem is $T$-tractable with parameter $\vp$ iff there exists a positive constant $C_{\vp}$ such that
\begin{equation} \label{eq:tract_def}
	\comp(\varepsilon,d) \le C_{\vp}\, T(\varepsilon^{-1},d,\vp) \qquad \forall \varepsilon >0, \ d \in \mathbb{N}.
\end{equation}
A problem is \emph{strongly}
$T$-tractable with parameter $\vp$ iff the complexity is independent of the dimension, that is, there exists a positive constant $C_{\vp}$ such that
\begin{equation} \label{eq:strong_tract_def}
	\comp(\varepsilon,d) \le C_{\vp}\, T(\varepsilon^{-1},1,\vp) \qquad \forall \varepsilon >0 , \ d \in \mathbb{N}.
\end{equation}

@@

If \eqref{eq:tract_def} holds for some $\vp$, it clearly holds for larger, $\vp$.  We are often interested in the optimal or smallest $\vp$ for which \eqref{eq:tract_def} holds.  Let $\cp_{\textup{true}} : = \{\vp^* : \eqref{eq:tract_def} \text{ holds }\forall \vp \in (\vp^*,\boldsymbol{\infty})\}$
The  set of optimal parameters, $\mathcal{P}_{\text{opt}}$ is defined as
\begin{equation}
	\mathcal{P}_{\textup{opt}} : = \{\vp^* \in \cp_{\textup{true}} :  \vp* \notin [\vp^\dagger,\boldsymbol{\infty}) \ \forall \vp^\dagger \in  \cp_{\textup{true}} \setminus \{\vp^*\} \}.
\end{equation}
In this article we allow $\vp$ to be a scalar or vector as the situation requires.

@@

In the sections below we prove necessary and sufficient conditions on tractability as generally defined in \eqref{eq:tract_def} and \eqref{eq:strong_tract_def}.  These conditions  that involve the boundedness of sums defined in terms of $\{T(\lambda_{i,d}^{-1},\cdot, \cdot)\}_{i=1}^\infty$.  In practice, it may be easier to verify whether or not these conditions hold than to deal with \eqref{eq:comp_def} directly.

@@

\begin{theorem}\label{thm_main_strong_tract2}
A problem is strongly $T$-tractable iff there exists $\vp \in [\vzero, \boldsymbol{\infty})$ and an integer $L(\vp) > 0$ such that
\begin{equation} \label{eq:strong_tractiff3}
     S(p):=\sup_{d \in \naturals} \sum_{i = L(\vp)}^\infty \frac{1}{T(\lambda_{i,d}^{-1},1,\vp)} < \infty.
\end{equation}
If \eqref{eq:strong_tractiff3} holds for some $\vp$, let  $\widetilde{\cp}_{\textup{true}} : = \{\vp^* : \eqref{eq:strong_tractiff3} \text{ holds }\forall \vp \in (\vp^*,\boldsymbol{\infty})\}$.  Then the set of optimal strong tractability parameters is
\[
	\cp_{\textup{opt}} = \widetilde{\mathcal{P}}_{\textup{opt}} :=
	\{\vp^* \in \widetilde{\cp}_{\textup{true}} :  \vp* \notin [\vp^\dagger,\boldsymbol{\infty}) \ \forall \vp^\dagger \in  \widetilde{\cp}_{\textup{true}} \setminus \{\vp^*\} \}.
\]
\end{theorem}

\begin{proof}
\textbf{Sufficient condition:}\\
Given that $T$ is increasing in its arguments, we have the following equivalent expression for the complexity:
    \begin{equation} \label{eq:comp_equiv}
    \comp(\varepsilon, d) &= \min \{n \in \natzero : \lambda_{n+1, d} \le \varepsilon\} = \min \biggl\{n \in \natzero : \frac{1}{T(\lambda_{n+1,d}^{-1},d,\vp)}\le \frac{1}{T(\varepsilon^{-1},d,\vp) }\biggr\}.
\end{equation}
We use this expression for $d=1$ here and for arbitrary $d$ in the next subsection

Also since the $\lambda_{i,d}$ are non-increasing, it follows that the $T(\lambda_{i,d}^{-1},1,\vp)$ are non-decreasing. In particular, for all $n\in \mathbb{N}$, we have

\begin{align*}
    \lambda_{n+1,d} \le \lambda_{n,d} \le \cdots \le \lambda_{1,d}
    & \implies \frac{1}{T(\lambda_{n+1, d}^{-1},1,\vp)} \le \frac{1}{T(\lambda_{n, d}^{-1},1,\vp)} \le \cdots \le \frac{1}{T(\lambda_{1, d}^{-1},1,\vp)} \\
    & \implies \frac{1}{T(\lambda_{n+1, d}^{-1},1,\vp) }
    \le \frac 1n \sum_{i=1}^n  \frac{1}{T(\lambda_{i, d}^{-1},1,\vp) }
    \le \frac 1n \sum_{i=1}^\infty  \frac{1}{T(\lambda_{i, d}^{-1},1,\vp)}.
\end{align*}
Thus, we can conclude from the previous line that
\begin{equation*}
    n \ge T(\varepsilon^{-1},1,\vp) \sum_{i=1}^\infty \frac{1}{T(\lambda_{i, d}^{-1},1,\vp)}
   \implies   \frac{1}{T(\lambda_{n+1, d}^{-1},1,\vp)} \le
   \frac 1n \sum_{i=1}^\infty \frac{1}{T(\lambda_{i, d}^{-1},1,\vp) } \le \frac{1}{T(\varepsilon^{-1},1,\vp)}.
\end{equation*}

This last inequality implies an upper bound on the complexity of
\begin{align*}
       \comp(\varepsilon,d)
       &\le 1 + T(\varepsilon^{-1},1,\vp) \sum_{i=1}^\infty \frac{1}{T(\lambda_{i, d}^{-1},1,\vp)} \\
       & = 1 + T(\varepsilon^{-1},1, \vp) \left [ \sum_{i=1}^{L(p)-1} \frac{1}{T(\lambda_{i, d}^{-1},1,\vp)}
       + \underbrace{\sum_{i=L(p)}^\infty \frac{1}{T(\lambda_{i, d}^{-1},1,\vp)}}_{\le S(p) <\infty\ \text{by \eqref{eq:strong_tractiff3}} }\right] \\
       & \le 1+ T(\varepsilon^{-1},1,\vp) \underbrace{\left[\frac{(L(p)-1)}{T(\lambda_{\max}^{-1},1,\vp)} + S(p)\right]}_{=:C_1(p)}\\
       & = 1+ T(\varepsilon^{-1},1,\vp) C_1 (p)\\
       &\le C_{\vp} T(\varepsilon^{-1},1,\vp),
\end{align*}
for some suitably chosen $C_{\vp}>0$. This means that we have strong $T$-tractability via \eqref{eq:strong_tract_def}, and verifies sufficiency of \eqref{eq:strong_tractiff3}.

\bigskip
\noindent \textbf{Necessary condition:} \\
Suppose that we have strong
$T$-tractability as defined in \eqref{eq:strong_tract_def}. That is, for some $\vp \ge \vzero$ there exists a positive constant $C_{\vp}$ such that
\[
\comp(\varepsilon,d)\le C_{\vp}\, T(\varepsilon^{-1},1,\vp)
\qquad \forall \varepsilon > 0, \ d \in \mathbb{N}.
\]
Since the eigenvalues $\lambda_{j,d}$ are non-increasing, we have
\begin{equation}\label{eq:lambda_K_strong3}
\lambda_{\lfloor C_{\vp}\, T(\varepsilon^{-1},1,\vp)\rfloor +1,d}\le \varepsilon.
\end{equation}

@@

For $\varepsilon$ taking on all positive values, $j(\varepsilon,\vp)$ takes on all values greater than or equal to $L(\vp)$, so
\[
j\le C_{\vp}\, T(\varepsilon^{-1},1,\vp)+1 \le C_{\vp} T(\lambda_{j,d}^{-1},1,\vp)+1, \qquad \forall j \ge L(\vp).
\]
This implies via our technical assumption \eqref{eq:ptauassume} that
\begin{equation*}
 K_{\vp,\tau}\,T (\lambda_{j,d}^{-1},1,\tau \vp) \ge
 [T(\lambda_{j,d}^{-1},1, \vp)]^\tau
 \ge
  \left[\frac{(j-1)}{C_{\vp}}\right]^\tau \qquad \forall j \ge L(\vp)
\end{equation*}
Summing both sides of the inequality yields and noting that $L(\vp) \le L(\tau\vp)$, it follows that
\begin{multline*}
\sup_{d\in\naturals} \sum_{j=L(\tau\vp)}^\infty \frac{1}{T (\lambda_{j,d}^{-1},1, \tau \vp)} \le
\sup_{d\in\naturals} \sum_{j=L(\vp)}^\infty \frac{1}{T (\lambda_{j,d}^{-1},1, \tau \vp)} \\
\le K_{\vp,\tau}\, C_{\vp}^\tau
\sum_{j=L(\vp)}^\infty \frac{1}{(j-1)^\tau} \le
K_{\vp,\tau}\, C_{\vp}^\tau\, \zeta (\tau) < \infty,
\end{multline*}
where $\zeta$ denotes the Riemann zeta function.
This yields \eqref{eq:strong_tractiff3} with $p$ replaced by $\tau p$, and so we see the necessity of $\eqref{eq:strong_tractiff3}$.

\bigskip
\noindent \textbf{Optimality:} \\
To complete the proof, we must show that $\cp_{\textup{opt}} = \widetilde{\cp}_{\textup{opt}}$.  We do this by proving that $\cp_{\textup{true}} = \widetilde{\cp}_{\textup{true}}$.  Then, since $\cp_{\textup{opt}}$ is defined in terms of $\cp_{\textup{opt}}$ in the same way as $\widetilde{\cp}_{\textup{opt}}$ is defined in terms of $\widetilde{\cp}_{\textup{opt}}$, it follows that $\cp_{\textup{opt}} = \widetilde{\cp}_{\textup{opt}}$.

If $\vp^* \in \cp_{\textup{true}}$, then \eqref{eq:strong_tract_def} holds for all $\vp \in (\vp^*,\vinfty)$.  By the argument to prove the necessary condition, it follows that \eqref{eq:strong_tractiff3} must also hold for all $\vp \in (\vp^*,\vinfty)$, so $\vp^* \in \widetilde{\cp}_{\textup{true}}$.  Conversely, if $\vp^* \in \widetilde{\cp}_{\textup{true}}$, then \eqref{eq:strong_tractiff3}  holds for all $\vp \in (\vp^*,\vinfty)$.  By the argument to prove the sufficient condition, it follows that \eqref{eq:strong_tract_def} must also hold for all $\vp \in (\vp^*,\vinfty)$, so $\vp^* \in \cp}_{\textup{true}}$.

\noindent This concludes the proof of the theorem.
